\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,graphicx,hyperref,xcolor}
\usepackage[numbers,sort&compress]{natbib}
\newcommand{\rhofunc}{\rho^{\mathrm{func}}}

\title{Silently Absorbed or Loudly Flagged:\\
Local Falsifiability Governs How Transformers Respond\\
to Planted Errors in Their Own Reasoning}
\author{Elyas Obbad \\ Stanford University \\ \texttt{eobbad@stanford.edu}
\and Brando Miranda \\ Stanford University}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Compounding-error arguments against autoregressive reasoning assume mid-trajectory
errors are unrecoverable. We test the premise causally: take multi-hop deduction
problems a pretrained transformer solves, inject a single controlled sentence into the
model's \emph{own} proof, remove all external feedback, and formally validate every
step of the continuation. Our central finding required catching our own construct
error: a perturbation generator that picks facts ``off the gold proof's path'' ---
the natural reading of \emph{wrong} --- produces statements that are \textbf{entailed-true
92\% of the time} in PrOntoQA worlds. Auditing injection truth status inverts the
apparent result. True-but-irrelevant interruptions are shrugged off (87\% formally
valid re-derivation, 0\% error adoption). Genuinely false injections split by
\emph{local falsifiability}: lies contradicting evidence available within one
inference hop are loudly flagged (up to 39\% verbalized doubt) and largely survived
(64--75\% valid re-derivation), while lies that can only be falsified by global
entailment checking are \textbf{silently absorbed} --- 0\% verbalized doubt, valid
re-derivation collapsing to 23--34\%, and the planted fact entering the model's
subsequent derivation in 29--60\% of runs. All falsehoods cause damage exceeding a
benign-paraphrase control ($p<.03$ at every injection position); a formal
step-validator separates genuine re-derivation from goal parroting, which inflates
naive final-answer ``recovery'' by 10--20 points. The gradient is parametric: placing the falsifying evidence exactly $k$ inference
hops away yields monotonically falling doubt ($.39 \to .32 \to .23 \to .00$) and
recovery as $k$ grows. Three further results bound the claim. Recovery survives a
question-format condition in which the queried statement's truth value is unknown
(57\% formally validated, so it is not goal parroting). The regime sets the price:
on GSM8K and chained arithmetic, where intermediates are load-bearing, absorption
reaches 88--100\% --- and the GSM8K case shows models re-check what they can re-read
in context, not what they must recompute. Scale attenuates absorption
(32B recovers from hard-to-check lies at $\sim$3$\times$ the 7B rate) but not
silence, and a second model family (OLMo-2) replicates the recovery gradient while
verbalizing nothing --- flagging is training-dependent, recovery is not. All rates are measured on a doubly
filtered cohort (instances the model solves, whose unperturbed proofs formally
validate) and are upper bounds. For the compounding debate the upshot is sharpened
rather than settled: per-step error \emph{absorption} is real precisely when errors
are hardest to check --- the dangerous case --- while cheaply-checkable errors are
not absorbed. Off-path $\neq$ false is a construct trap for any perturbation study
on entailment-dense tasks; we release the truth-status audit with the pipeline.
\end{abstract}

\section{Introduction}
A stylized argument pattern, assembled from ingredients in the compounding-error
literature \citep{lecun2022path,dziri2023faith}, holds that per-token commitment makes
long autoregressive generations collapse: if each step errs with probability
$\varepsilon$ and errors are independent, constant-rate, and unrecoverable, success
decays like $(1-\varepsilon)^T$. The \emph{unrecoverable} premise is the load-bearing
one, and it is usually attacked externally, via verifiers and tools. We ask the internal
question: when a single error is planted in a pretrained transformer's own reasoning
trajectory, with all external feedback removed, what does the model's subsequent
computation actually do with it?

Prior interventional work perturbs visible chain-of-thought and reads out behavioral
robustness \citep{vonrecum2026robust}; redundancy and self-repair results
\citep{mcgrath2023hydra,lad2024stages} suggest internal robustness mechanisms exist.
What is missing is (i) \emph{semantic control over what the perturbation actually is}
--- wrong relative to what? --- and (ii) a behavioral metric strong enough to
distinguish re-derivation from reaching a goal stated in the prompt. Both gaps turn out
to be load-bearing. Controlling them yields a result neither the fragility nor the
robustness literature predicts cleanly: the model's response to a planted error is
governed by how cheaply the error can be falsified against its local context.

\paragraph{Contributions.}
\textbf{(1)} A perturb-and-trace protocol on the model's own solved rollouts with a
\emph{formal step-validator}: each continuation is classified as valid re-derivation,
poisoned (a later step derivable only through the planted error), parroted (goal
reached via an underivable step), or derailed. Naive final-answer accuracy overstates
recovery by 10--20 points.
\textbf{(2)} A \emph{truth-status audit} exposing a construct trap: in entailment-dense
worlds, ``off the gold path'' injections are usually entailed-true (92\% here).
Conclusions about error recovery invert once injections are audited for actual falsity.
\textbf{(3)} Six perturbation families spanning a falsifiability gradient --- benign
paraphrase, true interruption, irrelevant rule, contradiction of an established fact,
negation of the imminent step, and unfalsifiable-without-global-checking false fact ---
each $n{=}130$ runs per injection position (35 for the last, availability-limited).
\textbf{(4)} The finding: \emph{silent absorption of hard-to-check lies}
(0\% doubt, 29--60\% poisoning) vs.\ \emph{loud flagging of cheap-to-check lies}
(18--39\% doubt, 0\% measurable poisoning, 64--75\% valid re-derivation), with
true interruptions shrugged off and all falsehoods significantly worse than the benign
control.

\section{Setup}
\paragraph{Task and model.} PrOntoQA-OOD \citep{saparov2023prontoqa} generates
multi-hop deductions over fictional ontologies in machine-parseable sentences. We use
500 instances (4--6 hops), few-shot prompting in the canonical proof style, and frozen
Qwen2.5-7B-Instruct (greedy decoding). The model solves 91\% of instances; the analysis
cohort is the solved subset whose gold rollout itself passes the formal validator
(84\%; $n{=}168$ instances, $n{=}130$ runs per family$\times$position cell).

\paragraph{Protocol.} We parse the model's own rollout into sentences, select an
intermediate derived fact at an early/mid/late position, replace it, and let the model
continue greedily with no feedback, no critique prompt, and a fixed budget. Hidden
states are cached at five depths for all rollouts. For each instance and position, a
null ensemble of $n{=}8$ temperature-0.8 continuations from the same prefix with the
\emph{correct} step (filtered to those remaining correct) calibrates all latent
comparisons.

\paragraph{Perturbation families and the falsifiability gradient.}
\textbf{(a) Benign paraphrase}: the correct step with a discourse marker prepended ---
semantically identical, controls for any-disruption effects.
\textbf{(b) True interruption}: an entity-category fact off the gold proof's path but
\emph{entailed by the world} (see \S\ref{sec:audit} --- this family was designed as
``wrong'' and discovered to be true).
\textbf{(c) Distractor}: a fabricated rule irrelevant to the entity.
\textbf{(d) Contradiction}: negation of the most recently \emph{established} fact ---
false, and in direct conflict with stated context (checkable at distance zero).
\textbf{(e) Negated step}: negation of the fact the model was about to derive ---
false, not conflicting with any stated sentence, but falsifiable by a one-hop
inference from context.
\textbf{(f) Unfalsifiable-locally false fact}: an entity-category assertion verified
non-entailed under the world's rule closure --- detectable only by global checking.
Constructions (d)--(f) are verified false (and (b)'s truth verified) against the rule
closure, per-injection.

\section{Metrics}
\paragraph{Formal validation.} A continuation is a \textbf{valid re-derivation} iff
every entity-fact sentence is entailed by the true world state (premises plus prior
valid derivations) under the transitive closure of the question's rules, and the final
statement matches the target. \textbf{Poisoned}: some step is derivable \emph{only}
through the planted atom. \textbf{Parroted}: target reached despite an underivable
intermediate step. \textbf{Derailed}: wrong final statement. Discourse markers are
stripped before parsing (a bug our benign control caught: the model adopts injected
style, which broke naive parsing); unparsed runs are excluded and counted.
Two scope notes, stated plainly: closure entailment permits step-skipping, so we
separately measured derivation-free goal-jumps (4.7--9.2\% of valid re-derivations at
early/mid positions; at late positions gold itself has no remaining intermediate
steps); and poisoning is mechanically measurable only for \emph{positive category}
injections --- the grammar derives nothing from negated facts --- so 0\% poisoning for
families (d)--(e) is a measurement bound, not evidence of resistance.

\paragraph{Doubt.} A lexical detector (wait/however/but/actually/contradict/\ldots)
over the continuation. Crude, but the cross-family contrasts it yields exceed
$5\sigma$.

\section{The truth-status audit}
\label{sec:audit}
Our original ``wrong-category'' generator guaranteed injections were off the gold
proof's path. Auditing every injection against the world's rule closure showed
\textbf{92\% were entailed-true}: PrOntoQA entities carry second premises and distractor
rule branches, so most off-path categories are still true of the entity. Splitting that
family's outcomes by audited truth status inverts the headline
(Table~\ref{tab:invert}): true injections are shrugged off; the genuinely false
minority is absorbed. Every conclusion below uses truth-audited families. We flag this
as a general trap: in any entailment-dense benchmark, ``not in the gold derivation''
is a property of the \emph{proof}, not of the \emph{world}.

\begin{table}[h]
\centering\small
\begin{tabular}{lcccc}
\toprule
Injection (audited) & $n$ & Valid re-deriv. & Poisoned & Parroted \\
\midrule
entailed-true  & 355 & .868 & .000 & .082 \\
genuinely false & 35 & .257 & .457 & .200 \\
\bottomrule
\end{tabular}
\caption{The same generator's injections, split by audited truth status (wrong-category
family, all positions pooled). The apparent ``recovery from lies'' was recovery from
\emph{true statements}.}
\label{tab:invert}
\end{table}

\section{Results}
\subsection{The falsifiability gradient}
Table~\ref{tab:main} is the main result. Reading top to bottom:
benign disruption costs 8--23 points of valid re-derivation relative to the
construction ceiling (unperturbed gold = 100\% by cohort definition) --- \emph{any}
mid-proof interruption has a price. True interruptions and contradictions of stated
facts cost no more than benign ($|z|<1.5$, n.s.). Genuine falsehoods cost
significantly more than benign at every position (negated step: $z{=}{-}2.18/{-}3.20/{-}3.64$,
$p{=}.029/.0014/.0003$) --- and split sharply by checkability:

\begin{table}[h]
\centering\small
\begin{tabular}{llcccc}
\toprule
Family & Inj. & Valid re-deriv.\ [95\% CI] & Poisoned & Parroted & Doubt \\
\midrule
benign paraphrase & early & 0.77 [0.69,0.83] & --- & 0.05 & 0.08 \\
benign paraphrase & mid & 0.82 [0.74,0.87] & --- & 0.05 & 0.16 \\
benign paraphrase & late & 0.92 [0.85,0.95] & --- & 0.02 & 0.05 \\
\midrule
true interruption & early & 0.82 [0.75,0.88] & 0.02 & 0.07 & 0.01 \\
true interruption & mid & 0.75 [0.67,0.82] & 0.07 & 0.12 & 0.03 \\
true interruption & late & 0.86 [0.79,0.91] & 0.03 & 0.09 & 0.02 \\
\midrule
distractor rule & early & 0.73 [0.65,0.80] & 0.00 & 0.10 & 0.02 \\
distractor rule & mid & 0.72 [0.63,0.79] & 0.00 & 0.14 & 0.03 \\
distractor rule & late & 0.76 [0.68,0.83] & 0.00 & 0.15 & 0.03 \\
\midrule
contradiction (dist.\ 0) & early & 0.73 [0.65,0.80] & 0.00 & 0.11 & \textbf{0.23} \\
contradiction (dist.\ 0) & mid & 0.79 [0.71,0.85] & 0.00 & 0.02 & \textbf{0.27} \\
contradiction (dist.\ 0) & late & 0.85 [0.78,0.90] & 0.01 & 0.02 & \textbf{0.18} \\
\midrule
false, 1-hop checkable & early & 0.65 [0.56,0.72] & 0.00$^{\dagger}$ & 0.09 & \textbf{0.39} \\
false, 1-hop checkable & mid & 0.64 [0.55,0.72] & 0.00$^{\dagger}$ & 0.05 & \textbf{0.25} \\
false, 1-hop checkable & late & 0.75 [0.67,0.81] & 0.00$^{\dagger}$ & 0.05 & \textbf{0.18} \\
\midrule
false, globally checkable only & early & 0.34 [0.21,0.51] & \textbf{0.29} & 0.20 & 0.00 \\
false, globally checkable only & mid & 0.23 [0.12,0.39] & \textbf{0.60} & 0.17 & 0.00 \\
false, globally checkable only & late & 0.29 [0.16,0.45] & \textbf{0.43} & 0.23 & 0.06 \\
\bottomrule
\end{tabular}
\caption{Formally validated outcomes, Qwen2.5-7B, $n{=}130$/cell (35 for the last
family, availability-limited to $\sim$30\% of instances; a within-subset control shows
the negated-step family performs identically on those instances, .676 vs .677, so the
gap is a construction effect, not instance selection). $^{\dagger}$Poisoning is not
measurable for negated injections (see Metrics). ``---'': not applicable (injection
true by construction).}
\label{tab:main}
\end{table}

\textbf{Silent absorption.} For falsehoods detectable only by global entailment
checking, the model verbalizes doubt in 0/70 early+mid runs, completes a
confident-looking proof, and incorporates the planted fact into its derivation in
29--60\% of runs. \textbf{Loud flagging.} For falsehoods refutable by a one-hop check
against context, verbalized doubt jumps to 18--39\% (the highest of any family;
contradiction-vs-true-interruption doubt contrast $z{=}5.5$, $p<10^{-7}$) and valid
re-derivation stays at 64--75\%. The model's error response is gated not by falsity
but by the \emph{cost of the check}. The negated-step-vs-benign deficits survive Holm
correction across the three injection positions (adjusted $p = .029/.0028/.0009$);
per our pre-registration, the original wrong/early cell is the designated primary
contrast and all other behavioral contrasts are exploratory.

\paragraph{The gradient is parametric: a dose--response in inferential distance.}
Negating the fact the model will derive $k$ hops downstream places the falsifying
evidence exactly $k$ inferences away. At early injection ($n{=}130$/cell for
$k\le 3$), verbalized doubt falls --- $.392$ ($k{=}1$) $\to .323$ ($k{=}2$) $\to
.231$ ($k{=}3$) --- then \emph{plateaus} at $k{=}4,5$ ($.229, .246$; $n{=}96, 69$
--- only longer chains admit distant targets, a selection caveat), with the
global-only construction far below the plateau ($.000$). Validated recovery follows
the same decline-then-plateau shape ($.646 \to .562 \to .531 \to .594 \to .594$).
The pre-registered monotone prediction is confirmed over $k{=}1$--$3$ and amended to
decline-then-plateau beyond. At 32B the curve shifts up and compresses: recovery
$.699$--$.753$ at every $k$, doubt $.237/.151/.161$ --- scale lifts the curve rather
than steepening it. Checkability is a quantity, not a category; its behavioral
effect saturates a few hops out.

\paragraph{Prompting does not fix the dangerous case.} A verification instruction
("before using any stated fact, verify it follows from the rules; if a statement
conflicts, point this out\ldots") was pre-registered to cut silent absorption by
$\ge$50\%. It does not: poisoning on globally-checkable falsehoods falls only
$\sim$17\% relative (pooled $.438 \to .362$, $n{=}105$/arm), and doubt on those lies
stays near zero. What the instruction \emph{does} do is nearly double verbalized
doubt on already-checkable lies ($.392 \to .585$ early) while \emph{reducing} their
validated recovery ($.646 \to .554$; mid $.638 \to .469$) --- the model spends its
budget second-guessing what it could already check, echoing the reasoning-trained
model's reactivity cost. Verbalization is cheap to induce; checking what is locally
unfalsifiable is not.

\paragraph{The doubt instrument is not a regex artifact.} A Qwen2.5-32B judge,
prompted to classify each mid-injection continuation for expressed doubt or
self-correction, agrees with the lexical detector on 76--100\% of runs per family and
\emph{sharpens} the contrast: judge-rated doubt is .285 (1-hop falsehood) and .246
(contradiction) vs.\ .015 (true interruption) and .000 (globally-checkable falsehood);
the lexical detector's largest deviation (benign paraphrase, .162 lexical vs.\ .092
judge) reflects regex false positives on discourse markers, in the direction that
\emph{understates} our contrast.

\subsection{The regime supplies recoverability: three tasks, one protocol}
The same mid-trajectory corruption protocol across task structures (7B):
\begin{center}\small
\begin{tabular}{lccc}
\toprule
Task & Recovery & Absorption & $n$ \\
\midrule
PrOntoQA (redundant derivations) & .53--.86 & .00--.60$^{*}$ & 130/cell \\
GSM8K (natural-language math) & .088 & .877 & 57 \\
Chained arithmetic (no redundancy) & .00--.02 & .99--1.00 & 99/cell \\
\bottomrule
\end{tabular}
\end{center}
($^{*}$by family; GSM8K cohort is format-filtered --- correct greedy solutions with
$\ge$3 explicit computations, 57/250 --- selecting multi-step explicit-arithmetic
problems; corruption at the middle computation.) Redundant tasks afford
re-derivation; tasks where each intermediate is load-bearing convert nearly every
absorbed error into a wrong answer. Notably, the corrupted GSM8K value \emph{is}
locally checkable --- one recomputation --- yet absorption dominates: the checking
that drives the PrOntoQA gradient operates over \emph{stated context facts},
not over recomputation. Models re-check what they can re-read, not what they must
re-derive numerically. ``Recoverability'' is a property of the task's derivation
structure and the check's modality, at least as much as of the model.

\subsection{Recovery is not goal-anchor re-reading}
Two ablations target the concern that recovery is just attention re-reading the
stated target. \textbf{(a) Question format}: rephrasing instances as ``True or false:
$s$?'' --- with $s$ the entailed target on half the instances and its negation on the
other half --- removes the parrotable goal (the statement is visible; its truth value
is not). Validated recovery (correct verdict $+$ formally sound post-injection chain)
remains 57\% under 1-hop falsehoods and 64\% under benign paraphrase
($n{=}35/36$; QA gold solve rate 55\%, so the cohort is smaller and harder).
Attenuated, not collapsed: goal-anchoring is not a sufficient explanation.
\textbf{(b) Attention knockout} on the goal-statement span collapses \emph{both} the
falsehood arm (.015) \emph{and the benign arm} (.046) --- the masking breaks the task
(without a visible target, proof termination is undefined), so it cannot isolate the
recovery mechanism; we report it as a negative design result. Its salvageable
secondary: goal-masked continuations remain chain-sound at identical rates for benign
and falsehood arms (.45 vs .42), i.e., goal access matters for derivation quality
generally, not differentially for error recovery.

\subsection{Latent trace}
A null-calibrated functional divergence ($\rhofunc_h$: Jensen--Shannon divergence
between perturbed and gold next-token distributions at sentence boundaries, normalized
by the correct-null ensemble's spread; LM-head computation verified exactly against
live logits; instance-specificity verified by a shuffled-gold control, 1.49$\times$)
shows a consistent shape at well-powered horizons (early/mid injections, $h\le 4$,
$n\ge 106$ per point): a large spike at the first post-injection boundary, decaying
toward but remaining above the null band ($\rhofunc_4 \approx 1.2$--$2.3$ across
families). We report this as descriptive only, for two reasons measured in our own
data. First, the null ensemble is temperature-sampled while rollouts are greedy, a
decoding mismatch that inflates the denominator by an unquantified factor. Second, and
decisively: the \emph{benign paraphrase} family shows the largest early-injection
divergence of any family ($\rhofunc_4{=}2.32$ vs.\ 1.9--2.2 for genuine falsehoods) ---
the model adopts the injected discourse style for the rest of its proof, and the
functional metric tracks that surface change as strongly as it tracks semantic damage.
As instrumented, $\rhofunc$ cannot distinguish a lie from a stylistic tic.
Three follow-up controls close the question. \emph{Matched decoding}: sampling
perturbed continuations at the null's temperature ($T{=}0.8$, $n{=}4$/site) removes
the greedy-vs-sampled confound, and benign $\approx$ falsehood persists (h1: 22.4 vs
16.3; h4: 1.36 vs 1.58) --- the null result is not a decoding artifact.
\emph{Ensemble stability}: across reasonable null configurations ($T \in
\{0.6, 0.8, 1.0\}$, $n \in \{4, 8, 16, 32\}$) the instance-level $\rhofunc$ median
ranges 1.9--7.9 with interquartile ranges spanning two orders of magnitude --- the
ratio is unstable at feasible ensemble sizes. Late-injection horizons beyond $h{=}2$
collapse to $n\le 12$ and support no conclusions. Raw hidden-state geometry tells the same story: null-calibrated cosine
divergence at the final layer is statistically indistinguishable between benign
paraphrase (median 2.57) and verified falsehoods (2.52--2.99) at well-powered
horizons --- surface form dominates content in geometric trajectory comparisons too.
(A layer-0 lexical control was attempted but is degenerate at sentence boundaries,
where the position is always the period token; a pooled-sentence control is future
work.) We accordingly make \emph{no} latent-recovery claims: the paper's evidence is
behavioral and formally validated.

\subsection{Scale attenuates absorption but not silence; the gradient generalizes
across model families}
Running the truth-audited falsehood families across Qwen2.5 sizes answers the scale
question --- and partially falsifies our pre-registered prediction (we predicted
absorption persists; it persists \emph{attenuated}):
\begin{center}\small
\begin{tabular}{lcccc}
\toprule
 & \multicolumn{3}{c}{globally-checkable falsehood} & 1-hop falsehood \\
Model & valid & poisoned & doubt & doubt \\
\midrule
Qwen2.5-1.5B & .15--.46 & .31--.62 & .000 & .000--.018 \\
Qwen2.5-7B   & .23--.34 & .29--.60 & .000--.057 & .18--.39 \\
Qwen2.5-32B  & .60--.73 & .10--.37 & .033--.067 & .13--.28 \\
\bottomrule
\end{tabular}
\end{center}
Recovery from hard-to-check lies roughly triples from 7B to 32B and poisoning roughly
halves ($n{=}30$--$35$/cell for this family) --- scale buys checking. But the
\emph{silence} is scale-invariant: at every size, absorbed lies go unverbalized.
Separately, the pre-audit ``wrong-category'' sweep (predominantly true interruptions)
shows robustness to true interruptions also scaling (.60--.67 $\to$ .75--.86 $\to$
.88--.96; parroting .11--.20 $\to$ .07--.12 $\to$ .01--.02).

A second model family, OLMo-2-7B-Instruct ($n{=}76$/cell; solve rate .69), replicates
the \emph{recovery} gradient (benign .72 $\ge$ 1-hop falsehood .53 at mid; poisoning
concentrated in globally-checkable falsehoods, .22--.33 at $n{=}9$) but \textbf{not
the flagging}: OLMo verbalizes essentially nothing in any family (.00--.026),
recovering from checkable lies as silently as Qwen-1.5B does. The behavioral
checkability gradient appears model-general; \emph{verbalization of detected errors
is model- and training-dependent} --- recovery does not require verbalization even
within a single model.

The training axis is isolated by DeepSeek-R1-Distill-Qwen-7B (same backbone,
RL reasoning training; injection into the visible proof body, think block stripped
for validation; 29\% solve, 8--16\% unparsed, reported per pre-registration).
Pre-registered prediction --- doubt rises, recovery flat --- confirmed: visible
flagging of 1-hop falsehoods reaches \textbf{.672} ($n{=}122$; vs.\ .25 Qwen-7B,
.00 OLMo) while validated recovery stays flat (.689 vs.\ .638), and the flagging is
content-selective (benign paraphrase: .008). The verbalization spectrum --- silent
(OLMo, Qwen-1.5B) $\to$ moderate (Qwen-7B/32B) $\to$ loud (R1) --- at roughly
constant recovery supports a clean separation: \emph{flagging is trained; recovery
is intrinsic}. (R1's private channel is uninformative here: 100\% of its gold think
blocks contain doubt vocabulary at baseline --- a ceiling. R1 also handles the benign
style perturbation \emph{worse} than its base model, .549 vs.\ .815: reasoning
training buys error verbalization at some cost in disruption robustness.)

\section{Limitations}
Task coverage now spans two synthetic families plus GSM8K (format-filtered cohort,
57/250 --- selection toward multi-step explicit arithmetic; Lean theorem proving is
the next arm); two base-model families (Qwen2.5, OLMo-2; a reasoning-tuned arm is in
progress), greedy gold rollouts. The Prove-format
prompt states the target; the validator classifies parroting separately and the
QA-format condition shows recovery survives without a parrotable goal, but
goal-directedness still aids derivation quality generically (knockout secondary
result). The globally-checkable-only falsehood family has $n{=}35$/cell and exists
only on $\sim$30\% of instances (a within-subset control rules out instance
selection); the QA condition has $n{=}35$--$36$/cell. ``Local falsifiability'' is
operationalized at two levels (distance 0--1 vs.\ global); the parametric
inferential-distance version is future work. Latent and geometric metrics are
reported as negative/descriptive results (\S5.2). Closure-based validity permits
step-skipping (quantified at 4.7--9.2\%). One seed of null sampling; behavioral
results use deterministic decoding throughout.

\section{Related work}
\textbf{CoT interventions and error recovery.} \citet{vonrecum2026robust} perturb
visible CoT in reasoning-tuned models and find behavioral robustness with
early-intervention degradation and doubt/length effects; our truth-status audit
suggests such interventions need per-injection falsity verification for
``adversarial'' conditions to mean what they claim. \citet{yee2024dissociation}
precede our re-derivation/parroting split with a faithful/unfaithful recovery
dichotomy via manual annotation; our delta is automated formal validation, semantic
(vs.\ numerical) error control, the truth-status audit, and the falsifiability
gradient. Their finding that recovery is more frequent for \emph{obvious} errors
independently anticipates checkability-gating. \citet{lanham2023measuring} introduced
adding-mistakes interventions for faithfulness; \citet{turpin2023unfaithful} showed
stated reasoning can omit operative causes --- consistent with silent absorption.
\citet{zhang2023snowball} showed models commit to early mistakes they can separately
recognize; our results locate when that recognition is and is not deployed in-rollout.
\citet{huang2024cannot} found intrinsic self-correction weak at the multi-turn level;
silent absorption is its single-pass analogue, while flagging shows the capacity
exists when the check is local.
\textbf{Self-repair and redundancy.} \citet{mcgrath2023hydra,lad2024stages} document
structural self-repair; silent absorption is the complementary failure: robustness
machinery that smooths over content errors it cannot cheaply check.
\textbf{Trajectory geometry.} \citet{sun2026trajectories} characterize reasoning as
step-specific representational trajectories with correctness signals; our
style-propagation result (benign $\approx$ falsehood in both functional and geometric
divergence) is a caution for cross-trace comparisons in that line.
\textbf{Compounding.} \citet{dziri2023faith,lecun2022path} supply the ingredients of
the compounding frame our results sharpen: per-step absorption is real exactly when
errors are locally unfalsifiable, and task derivation structure (\S5.3) determines
the price.

\section*{Reproducibility}
Single 8$\times$H200 node, $\sim$8 GPU-hours total. Pre-registered plan
(\texttt{PLAN.md}, written before the first run), all scripts, per-run JSONL logs,
truth-status audit, and result JSONs accompany the paper
(\texttt{experiments/09\_latent\_recovery/}).

\bibliographystyle{plainnat}
\begin{thebibliography}{9}
\bibitem[Dziri et al.(2023)]{dziri2023faith} N. Dziri et al.
\newblock Faith and Fate: Limits of Transformers on Compositionality.
\newblock \emph{NeurIPS}, 2023. arXiv:2305.18654.
\bibitem[Lad et al.(2024)]{lad2024stages} V. Lad, J.H. Lee, W. Gurnee, M. Tegmark.
\newblock The Remarkable Robustness of LLMs: Stages of Inference?
\newblock arXiv:2406.19384, 2024.
\bibitem[LeCun(2022)]{lecun2022path} Y. LeCun.
\newblock A Path Towards Autonomous Machine Intelligence. OpenReview, 2022.
\bibitem[McGrath et al.(2023)]{mcgrath2023hydra} T. McGrath et al.
\newblock The Hydra Effect: Emergent Self-repair in Language Model Computations.
\newblock arXiv:2307.15771, 2023.
\bibitem[Saparov and He(2023)]{saparov2023prontoqa} A. Saparov and H. He.
\newblock Language Models Are Greedy Reasoners: A Systematic Formal Analysis of
Chain-of-Thought. \emph{ICLR}, 2023. arXiv:2210.01240.
\bibitem[Yee et al.(2024)]{yee2024dissociation} E. Yee, A. Li, C. Tang, Y.H. Jung,
R. Paturi, L. Bergen.
\newblock Dissociation of Faithful and Unfaithful Reasoning in LLMs.
\newblock arXiv:2405.15092, 2024.
\bibitem[Lanham et al.(2023)]{lanham2023measuring} T. Lanham et al.
\newblock Measuring Faithfulness in Chain-of-Thought Reasoning.
\newblock arXiv:2307.13702, 2023.
\bibitem[Turpin et al.(2023)]{turpin2023unfaithful} M. Turpin, J. Michael,
E. Perez, S. Bowman.
\newblock Language Models Don't Always Say What They Think.
\newblock \emph{NeurIPS}, 2023. arXiv:2305.04388.
\bibitem[Zhang et al.(2023)]{zhang2023snowball} M. Zhang, O. Press, W. Merrill,
A. Liu, N.A. Smith.
\newblock How Language Model Hallucinations Can Snowball.
\newblock arXiv:2305.13534, 2023.
\bibitem[Huang et al.(2024)]{huang2024cannot} J. Huang et al.
\newblock Large Language Models Cannot Self-Correct Reasoning Yet.
\newblock \emph{ICLR}, 2024. arXiv:2310.01798.
\bibitem[Sun et al.(2026)]{sun2026trajectories} L. Sun, H. Dong, B. Qiao, Q. Lin,
D. Zhang, S. Rajmohan.
\newblock LLM Reasoning as Trajectories: Step-Specific Representation Geometry and
Correctness Signals.
\newblock arXiv:2604.05655, 2026.
\bibitem[von Recum et al.(2026)]{vonrecum2026robust} A. von Recum, L. Girrbach,
Z. Akata.
\newblock Are Reasoning LLMs Robust to Interventions on Their Chain-of-Thought?
\newblock arXiv:2602.07470, 2026.
\end{thebibliography}

\end{document}
